
Reproducible Research Assignment 
======================================================


## Loading and preprocessing the data

Unzipping the data and reading it in as a csv

```{r}
if (!file.exists('activity.csv')) {
  unzip(zipfile = "activity.zip")
}

activityData <- read.csv(file="activity.csv", header=TRUE)

```

## What is the  mean total number of steps taken per day?

```{r}
# Calculate the total steps taken per day
totalSteps <- aggregate(steps ~ date, activityData, FUN=sum)

# Make a histogram of the total number of steps taken per day
hist(totalSteps$steps,
     main = "Total Steps per Day",
     xlab = "Number of Steps",
     col= "brown")

# Calculate and report the mean and median of total steps taken per day
meanSteps <- mean(totalSteps$steps, na.rm = TRUE)
medSteps <- median(totalSteps$steps, na.rm = TRUE)

```
The mean of total steps taken per day is `r meanSteps`
and the  median of total steps taken per day is `r medSteps`


## What is the average daily activity pattern?
```{r}
# Make a time series plot
library(ggplot2)
meanStepsByInt <- aggregate(steps ~ interval, activityData, mean)
ggplot(data = meanStepsByInt, aes(x = interval, y = steps)) +
  geom_line(col = "red") +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))

# Which 5-minute interval across all days contain the maximum number of steps
maxInt <- meanStepsByInt[which.max(meanStepsByInt$steps),]

```
The 5-minute interval across all day containing maximum steps is `r maxInt`

## Imputing missing values

```{r}
# Calculate and report the total number of missing values in the dataset
missingVals <- is.na(activityData$steps)
```

There are `r sum(missingVals)` missing values. I will replace these missing values with the 5-day mean of the interval.

```{r}
# Create a new dataset that is equal to the original dataset but with 
# the missing data filled in.
imp_activityData <- transform(activityData,
                              steps = ifelse(is.na(activityData$steps),
                                             meanStepsByInt$steps[match(activityData$interval, 
                                                                        meanStepsByInt$interval)],
                                             activityData$steps))

# Make a histogram of the total number of steps taken each day and
# and report the mean and median.
impStepsByInt <- aggregate(steps ~ date, imp_activityData, FUN=sum)
hist(impStepsByInt$steps,
     main = "Imputed Number of Steps Per Day",
     xlab = "Number of Steps")

impMeanSteps <- mean(impStepsByInt$steps, na.rm = TRUE)
impMedSteps <- median(impStepsByInt$steps, na.rm = TRUE)
diffMean = impMeanSteps - meanSteps
diffMed = impMedSteps - medSteps
diffTotal = sum(impStepsByInt$steps) - sum(totalSteps$steps)
```
There is a difference  of`r diffMean` in the mean steps of the two data. There is a difference of 
`r diffMed` in the median of the data sets and there is a difference of `r diffTotal` in the total of
the two data sets.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Create a new factor variable in the dataset with two levels - "weekend" and "weekday"
DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday'))
      return ("weekeday")
  else if (day %in% c('Saturday', 'Sunday'))
      return ("weekend")
  else
      stop ("Invalid Date Format.")
}
imp_activityData$date <- as.Date(imp_activityData$date)
imp_activityData$day <- sapply(imp_activityData$date, FUN = DayType)

# Make a panel plot containnig a time-series plot of the 5-minute interval
# and the average number of steps taken across all weekdays or weekends
meanStepsByDay <- aggregate(steps ~ interval + day, imp_activityData, mean)
ggplot(data = meanStepsByDay, aes(x = interval, y = steps)) + 
  geom_line(col="blue") +
  facet_grid(day ~ .) +
  ggtitle("Average Daily Activity Pattern") +
  xlab("5-minute Interval") +
  ylab("Average Number of Steps") +
  theme(plot.title = element_text(hjust = 0.5))
```